---
title: "Full Study Data Analysis"
date: '2023-09-07'
output: 
  html_document:
    toc: true
    toc_depth: 5
    number_sections: true
    toc_float: true
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(ggplot2)
library(tidyverse)
```

Load data
```{r}

rating <- read.csv('data/ratings_with_change.csv')
applicant <- read.csv('data/applications.csv')
IAT <- read.csv('data/IAT_score.csv')
time <- read.csv('data/focus_time_with_change.csv')

```


Combine datasets 
```{r}
df <- merge(rating, applicant, by.x='applicant_id', by.y='id')

df <- merge(df, time, by=c("user_id", "applicant_id"))

df <- merge(df, IAT, by="user_id")

```


Check distribution
```{r}
t1 <- df %>%
  group_by(user_id) %>%
  summarize(min = (min(total_before)),
            max = (max(total_before)),
            mean = mean(total_before))
summary(t1)
```
```{r}
t1 <- t1 %>%
  summarise(mean_min=mean(min),
            mean_max=mean(max),
            mean_mean=mean(mean))
t1
```


```{r}
t2 <- df %>%
  group_by(user_id) %>%
  summarize(min = (min(rating_before)),
            max = (max(rating_before)),
            mean = mean(rating_before))
summary(t2)
```


```{r}
t2 <- t2 %>%
  summarise(mean_min=mean(min),
            mean_max=mean(max),
            mean_mean=mean(mean))
t2
```



# Data process: normalize by participant

Define normalization function
```{r}
norm <- function(x) {
  min <- min(x)
  max <- max(x)
  r <- max-min 
  norm_x <- 0
  if (r==0) {
    # if all numbers in x are 0, norm_x = 0
  } 
  else {
    #else norm_x = (x - min)/(max - min)
    for (i in 1:length(x)) {
      norm_x[i]<- (x[i] - min)/r
    }
  }
  return(norm_x)
}
```


## Rating
```{r}
for (i in unique(df$user_id)) {
  df[df$user_id==i, 'norm_rating_before'] <- norm(df[df$user_id==i, ]$rating_before)
  df[df$user_id==i, 'norm_rating_after'] <- norm(df[df$user_id==i, ]$rating_after)
  df[df$user_id==i, 'norm_rating_change'] <- norm(df[df$user_id==i, ]$rating_change)
}
```


## Total Focus Time
```{r}
for (i in unique(df$user_id)) {
  df[df$user_id==i, 'norm_total_before'] <- norm(df[df$user_id==i, ]$total_before)
  df[df$user_id==i, 'norm_total_after'] <- norm(df[df$user_id==i, ]$total_after)
  df[df$user_id==i, 'norm_total_change'] <- norm(df[df$user_id==i, ]$total_change)
}
```



```{r}
write.csv(df,'full_study_processed.csv')
```


# H1
Applicants’ gender and race will impact their evaluations


```{r}
h1 <- df
```


## H1.1
Female and non-binary applicants will get less review time than male applicants.\

### Delta Mean Time
Mean focus time (after normalization) by gender:\
Female	0.2870896				\ 
Male	0.3657257			\ 
Non-binary	0.3812948	\	


Delta mean time (M-F) is 0.07863, p-value=0.00167\
Delta mean time (M-NB) is -0.01556, p-value=0.5473

```{r}
r1<- h1 %>%
  group_by(GENDER) %>%
  summarise(aver_time = mean(norm_total_before))
r1

print(paste0('Delta mean time (M-F) is ', r1[2,2]-r1[1,2]))
print(paste0('Delta mean time (M-NB) is ', r1[2,2]-r1[3,2]))
```

### t-test
```{r}
# T-test: whether true difference exists

# Male & Female
t.test(h1[h1$GENDER=='Male',]$norm_total_before, h1[h1$GENDER=='Female',]$norm_total_before, var.equal=FALSE)
# Male & non-binary
t.test(h1[h1$GENDER=='Male',]$norm_total_before, h1[h1$GENDER=='Non-binary',]$norm_total_before, var.equal=FALSE)

```


### Graph

Bootstrapping CI

```{r}
set.seed(10)
M <- 10000
# Empty matrix to save the values
delta_m <- matrix(NA, M)
delta_f <- matrix(NA, M)
delta_nb <- matrix(NA, M)

for(i in 1:M){
  resample <- h1[sample(nrow(h1), replace = T),]
  boots_time_m <- mean(resample[resample$GENDER=='Male',]$norm_total_before)
  boots_time_f <- mean(resample[resample$GENDER=='Female',]$norm_total_before)
  boots_time_nb <- mean(resample[resample$GENDER=='Non-binary',]$norm_total_before)
  delta_m[i] <- boots_time_m
  delta_f[i] <- boots_time_f
  delta_nb[i] <- boots_time_nb
}

se_m <- sd(delta_m)
se_f <- sd(delta_f)
se_nb <- sd(delta_nb)

q_val <- qt(0.95, M-1)
r1[r1$GENDER == 'Male', 'low_CI'] <- mean(delta_m) - q_val*se_m
r1[r1$GENDER == 'Female', 'low_CI'] <- mean(delta_f) - q_val*se_f
r1[r1$GENDER == 'Non-binary', 'low_CI'] <- mean(delta_nb) - q_val*se_nb

r1[r1$GENDER == 'Male', 'high_CI'] <- mean(delta_m) + q_val*se_m
r1[r1$GENDER == 'Female', 'high_CI'] <- mean(delta_f) + q_val*se_f
r1[r1$GENDER == 'Non-binary', 'high_CI'] <- mean(delta_nb) + q_val*se_nb

r1
```



```{r}
h11<-ggplot()+
  geom_point(data=r1, aes(y=GENDER, x=aver_time, color=GENDER), size=3)+
  geom_text(data=r1, aes(label=format(round(aver_time, 3), nsmall=3), 
                         x=aver_time, y=GENDER),hjust=0.5, vjust=-1)+
  geom_errorbar(data=r1, aes(x=aver_time,y=GENDER, xmin = low_CI, xmax = high_CI, 
                             color=GENDER), width=0.3)+
  geom_point(data=h1, aes(y=GENDER, x=norm_total_before, color=GENDER),
             size=1, alpha=0.1)+
  labs(y="Gender", x="Average Review Time")+
  theme_bw()+
  scale_color_manual(values = c("#00AFBB", "#E7B800", "#FC4E07"))+
  theme(axis.text.x=element_text(size=12),
        axis.text.y=element_text(size=12),
        legend.text=element_text(size=12),
        text = element_text(size = 14),legend.position="top")
h11
ggsave(file="H1.1.pdf", width=5, height=3.5)

```


## H1.2. 
Black applicants will get less review time than white applicants. \

### Delta Mean Time

Mean focus time (after normalization) by race:\
Black	0.3552549			\
White	0.3340815		\

Delta mean time (W-B) is -0.02117, p-value=0.3047.

```{r}
r2<- h1 %>%
  group_by(RACE) %>%
  summarise(aver_time = mean(norm_total_before))
r2

print(paste0('Delta mean time (W-B) is ', r2[2,2]-r2[1,2]))
```


### t-test
```{r}
# T-test: true difference exists

t.test(h1[h1$RACE=='Black',]$norm_total_before, h1[h1$RACE=='White',]$norm_total_before, var.equal=FALSE)

```




### Graph

Bootstrapping CI

```{r}
set.seed(10)
M <- 10000
# Empty matrix to save the values
delta_b <- matrix(NA, M)
delta_w <- matrix(NA, M)

for(i in 1:M){
  resample <- h1[sample(nrow(h1), replace = T),]
  boots_time_w <- mean(resample[resample$RACE=='White',]$norm_total_before)
  boots_time_b <- mean(resample[resample$RACE=='Black',]$norm_total_before)
  delta_w[i] <- boots_time_w
  delta_b[i] <- boots_time_b
}

se_w <- sd(delta_w)
se_b <- sd(delta_b)

q_val <- qt(0.95, M-1)
r2[r2$RACE == 'White', 'low_CI'] <- mean(delta_w) - q_val*se_w
r2[r2$RACE == 'Black', 'low_CI'] <- mean(delta_b) - q_val*se_b

r2[r2$RACE == 'White', 'high_CI'] <- mean(delta_w) + q_val*se_w
r2[r2$RACE == 'Black', 'high_CI'] <- mean(delta_b) + q_val*se_b

r2
```

```{r}
h12<-ggplot()+
  geom_point(data=r2, aes(y=RACE, x=aver_time, color=RACE), size=3)+
  geom_text(data=r2, aes(label=format(round(aver_time, 3), nsmall=3), 
                         x=aver_time, y=RACE),hjust=0.5, vjust=-1)+
  geom_errorbar(data=r2, aes(x=aver_time,y=RACE, xmin = low_CI, xmax = high_CI, 
                             color=RACE), width=0.3)+
  geom_point(data=h1, aes(y=RACE, x=norm_total_before, color=RACE),
             size=1, alpha=0.1)+
  labs(y="Race", x="Average Review Time")+
  theme_bw()+
  scale_color_manual(values = c("#52854C", "#4E84C4"))+
  theme(axis.text.x=element_text(size=12),
        axis.text.y=element_text(size=12),
        legend.text=element_text(size=12),
        text = element_text(size = 14),legend.position="top")
h12
ggsave(file="H1.2.pdf", width=4.5, height=3)

```




## H1.3
Female and non-binary applicants will be rated lower than their equally-competent male counterparts.\

### Delta Mean Rating
Meaning rating (after normalization) by gender:\
Female	0.5236511				\
Male	0.541660			\
Non-binary	0.5335028	\
\
Delta mean rating (M-F) is 0.01800, p-value=0.5394\
Delta mean rating (M-NB) is 0.00815, p-value=0.7836

```{r}
# calculate the mean ratings by gender
r3<- h1 %>%
  group_by(GENDER) %>%
  summarise(aver_rating = mean(norm_rating_before))
r3

print(paste0('Delta mean rating (M-F) is ', r3[2,2]-r3[1,2]))
print(paste0('Delta mean rating (M-NB) is ', r3[2,2]-r3[3,2]))
```

### t-test
```{r}
# T-test: whether true difference exists

# Male & Female
t.test(h1[h1$GENDER=='Male',]$norm_rating_before, h1[h1$GENDER=='Female',]$norm_rating_before, var.equal=FALSE)
# Male & non-binary
t.test(h1[h1$GENDER=='Male',]$norm_rating_before, h1[h1$GENDER=='Non-binary',]$norm_rating_before, var.equal=FALSE)

```



### Graph

Bootstrapping CI

```{r}
set.seed(10)
M <- 10000
# Empty matrix to save the values
delta_m <- matrix(NA, M)
delta_f <- matrix(NA, M)
delta_nb <- matrix(NA, M)

for(i in 1:M){
  resample <- h1[sample(nrow(h1), replace = T),]
  boots_ratings_m <- mean(resample[resample$GENDER=='Male',]$norm_rating_before)
  boots_ratings_f <- mean(resample[resample$GENDER=='Female',]$norm_rating_before)
  boots_ratings_nb <- mean(resample[resample$GENDER=='Non-binary',]$norm_rating_before)
  delta_m[i] <- boots_ratings_m
  delta_f[i] <- boots_ratings_f
  delta_nb[i] <- boots_ratings_nb
}

se_m <- sd(delta_m)
se_f <- sd(delta_f)
se_nb <- sd(delta_nb)

q_val <- qt(0.95, M-1)
r3[r3$GENDER == 'Male', 'low_CI'] <- mean(delta_m) - q_val*se_m
r3[r3$GENDER == 'Female', 'low_CI'] <- mean(delta_f) - q_val*se_f
r3[r3$GENDER == 'Non-binary', 'low_CI'] <- mean(delta_nb) - q_val*se_nb

r3[r3$GENDER == 'Male', 'high_CI'] <- mean(delta_m) + q_val*se_m
r3[r3$GENDER == 'Female', 'high_CI'] <- mean(delta_f) + q_val*se_f
r3[r3$GENDER == 'Non-binary', 'high_CI'] <- mean(delta_nb) + q_val*se_nb

r3
```

```{r}
h13<-ggplot()+
  geom_point(data=r3, aes(y=GENDER, x=aver_rating, color=GENDER), size=3)+
  geom_text(data=r3, aes(label=format(round(aver_rating, 3), nsmall=3), 
                         x=aver_rating, y=GENDER),hjust=0.5, vjust=-1)+
  geom_errorbar(data=r3, aes(x=aver_rating,y=GENDER, xmin = low_CI, xmax = high_CI, 
                             color=GENDER), width=0.3)+
  geom_point(data=h1, aes(y=GENDER, x=norm_rating_before, color=GENDER),
             size=1, alpha=0.1)+
  labs(y="Gender", x="Average Competitiveness Rating")+
  scale_color_manual(values = c("#00AFBB", "#E7B800", "#FC4E07"))+
  theme_bw()+
  theme(axis.text.x=element_text(size=12),
        axis.text.y=element_text(size=12),
        legend.text=element_text(size=12),
        text = element_text(size = 14),legend.position="top")
h13
ggsave(file="H1.3.pdf", width=4.5, height=3.5)

```


## H1.4

Black applicants will be rated lower than their equally-competent white counterparts. \

### Delta mean rating
Meaning rating by race:\
Black	0.5411733			\
White	0.5246655	

\
Delta mean rating (W-B) is -0.01651, p-value= 0.4931\


```{r}
# calculate the mean ratings by gender
r4<- h1 %>%
  group_by(RACE) %>%
  summarise(aver_rating = mean(norm_rating_before))
r4
print(paste0('Delta mean rating (W-B) is ', r4[2,2]-r4[1,2]))
```

### t-test
```{r}
# T-test: true difference exists

t.test(h1[h1$RACE=='Black',]$norm_rating_before, h1[h1$RACE=='White',]$norm_rating_before, var.equal=FALSE)

```


### Graph

Bootstrapping CI

```{r}
set.seed(10)
M <- 10000
# Empty matrix to save the values
delta_b <- matrix(NA, M)
delta_w <- matrix(NA, M)

for(i in 1:M){
  resample <- h1[sample(nrow(h1), replace = T),]
  boots_ratings_b <- mean(resample[resample$RACE=='Black',]$norm_rating_before)
  boots_ratings_w <- mean(resample[resample$RACE=='White',]$norm_rating_before)
  delta_b[i] <- boots_ratings_b
  delta_w[i] <- boots_ratings_w
}

se_b <- sd(delta_b)
se_w <- sd(delta_w)

q_val <- qt(0.95, M-1)
r4[r4$RACE == 'White', 'low_CI'] <- mean(delta_w) - q_val*se_w
r4[r4$RACE == 'Black', 'low_CI'] <- mean(delta_b) - q_val*se_b

r4[r4$RACE == 'White', 'high_CI'] <- mean(delta_w) + q_val*se_w
r4[r4$RACE == 'Black', 'high_CI'] <- mean(delta_b) + q_val*se_b

r4
```

```{r}
h14<-ggplot()+
  geom_point(data=r4, aes(y=RACE, x=aver_rating, color=RACE), size=3)+
  geom_text(data=r4, aes(label=format(round(aver_rating, 3), nsmall=3), 
                         x=aver_rating, y=RACE),hjust=0.5, vjust=-1)+
  geom_errorbar(data=r4, aes(x=aver_rating,y=RACE, xmin = low_CI, xmax = high_CI, 
                             color=RACE), width=0.3)+
  geom_point(data=h1, aes(y=RACE, x=norm_rating_before, color=RACE),
             size=1, alpha=0.1)+
  labs(y="Race", x="Average Competitiveness Rating")+
  theme_bw()+
  scale_color_manual(values = c("#52854C", "#4E84C4"))+
  theme(axis.text.x=element_text(size=12),
        axis.text.y=element_text(size=12),
        legend.text=element_text(size=12),
        text = element_text(size = 14),legend.position="top")
h14
ggsave(file="H1.4.pdf", width=4.5, height=3)

```

```{r}
library(ggpubr)
ggarrange(ggarrange(h11, h12,nrow=2, align = "v"), ggarrange(h13, h14,nrow=2,align = "v"),
           ncol = 2)
ggsave(file="H1_combined.pdf", width=10.5, height=5.5)
```



# H2
Participants' review behavior (time spent) and decisions (ratings) will correlate with their implicit biases.

```{r}
h2 <- df
```



## H2.1 
Participants with higher gender IAT scores will spend less time on female and non-binary applicants than male applicants. \

New variable defined as:\
`gender_time_diff1` = mean time on Male applicant - mean time on Female applicant\
`gender_time_diff2` = mean time on Male applicant - mean time on Non-binary applicant\
\

Result: \
- M & F: For linear regression gender_time_diff1 ~ IAT_Score_Gender, slope = -0.06915, p-value=4.52e-05. Higher IAT score, more equally time spent between male & female applicants\

- M & N-B: For linear regression gender_time_diff2 ~ IAT_Score_Gender, slope = 0.06450, p-value = 0.000404 Higher IAT score, spend more time on male applicants than on non-binary applicants

```{r}
c1 <- h2 %>%
  group_by(user_id, GENDER) %>%
  summarise(gender_mean_time=mean(norm_total_before)) 


for (i in 1:length(h2$user_id)) {
  user = h2$user_id[i]
  h2[h2$user_id==user, 'gender_time_diff1'] <- c1[c1$user_id==user & c1$GENDER=='Male', ]$gender_mean_time - c1[c1$user_id==user & c1$GENDER=='Female', ]$gender_mean_time
  h2[h2$user_id==user, 'gender_time_diff2'] <- c1[c1$user_id==user & c1$GENDER=='Male', ]$gender_mean_time - c1[c1$user_id==user & c1$GENDER=='Non-binary', ]$gender_mean_time
}
```

### Male & Female

```{r}
mod.1 <- lm(gender_time_diff1 ~ IAT_Score_Gender, h2, na.action=na.omit)
summary(mod.1)
```

```{r}
h2 %>%
  ggplot(aes(x=IAT_Score_Gender,y=gender_time_diff1))+
  geom_point()+
  geom_smooth(method='lm')+
  theme_bw()
```

### Male & Non-binary

```{r}
mod.2 <- lm(gender_time_diff2 ~ IAT_Score_Gender, h2, na.action=na.omit)
summary(mod.2)
```
```{r}
h2 %>%
  ggplot(aes(x=IAT_Score_Gender,y=gender_time_diff2))+
  geom_point()+
  geom_smooth(method='lm')+
  theme_bw()
```

## H2.2 
Participants with higher race IAT scores will spend less time on black applicants than white applicants. \

New variable defined as:\
`race_time_diff` = mean time on White applicant - mean time on Black applicant\
\

Result: \
- For linear regression race_time_diff ~ IAT_Score_Race, slope = 0.117167, p-value=3.20e-15. Higher IAT score, spend more time on white applicants than on Black applicants\

```{r}
c2 <- h2 %>%
  group_by(user_id, RACE) %>%
  summarise(race_mean_time=mean(norm_total_before)) 


for (i in 1:length(h2$user_id)) {
  user = h2$user_id[i]
  h2[h2$user_id==user, 'race_time_diff'] <- c2[c2$user_id==user & c2$RACE=='White', ]$race_mean_time - c2[c2$user_id==user & c2$RACE=='Black', ]$race_mean_time
}
```


```{r}
mod.3 <- lm(race_time_diff ~ IAT_Score_Race, h2, na.action=na.omit)
summary(mod.3)
```

```{r}
h2 %>%
  ggplot(aes(x=IAT_Score_Race,y=race_time_diff))+
  geom_point()+
  geom_smooth(method='lm')+
  theme_bw()
```


## H2.3 
Participants with higher gender IAT scores will give lower ratings to female and non-binary applicants than male applicants. 

New variable defined as:\
`gender_rating_diff1` = mean rating for Male applicant - mean rating for Female applicant\
`gender_rating_diff2` = mean rating for Male applicant - mean rating for Non-binary applicant\
\

Result: \

- M & F: For linear regression gender_rating_diff1 ~ IAT_Score_Gender, slope = -0.060320, p-value=1.38e-05. Higher IAT score, give higher ratings to female applicants than male applicants\

- M & N-B: For linear regression gender_rating_diff2 ~ IAT_Score_Gender, slope = -0.033554, p-value = 0.002774. Higher IAT score, give higher ratings to non-binary applicants than male applicants

```{r}
c3 <- h2 %>%
  group_by(user_id, GENDER) %>%
  summarise(gender_mean_rating=mean(norm_rating_before)) 


for (i in 1:length(h2$user_id)) {
  user = h2$user_id[i]
  h2[h2$user_id==user, 'gender_rating_diff1'] <- c3[c3$user_id==user & c3$GENDER=='Male', ]$gender_mean_rating - c3[c3$user_id==user & c3$GENDER=='Female', ]$gender_mean_rating
  h2[h2$user_id==user, 'gender_rating_diff2'] <- c3[c3$user_id==user & c3$GENDER=='Male', ]$gender_mean_rating - c3[c3$user_id==user & c3$GENDER=='Non-binary', ]$gender_mean_rating
}
```

### Male & Female

```{r}
mod.4 <- lm(gender_rating_diff1 ~ IAT_Score_Gender, h2, na.action=na.omit)
summary(mod.4)
```

```{r}
h2 %>%
  ggplot(aes(x=IAT_Score_Gender,y=gender_rating_diff1))+
  geom_point()+
  geom_smooth(method='lm')+
  theme_bw()
```

### Male & Non-binary

```{r}
mod.5 <- lm(gender_rating_diff2 ~ IAT_Score_Gender, h2, na.action=na.omit)
summary(mod.5)
```


```{r}
h2 %>%
  ggplot(aes(x=IAT_Score_Gender,y=gender_rating_diff2))+
  geom_point()+
  geom_smooth(method='lm')+
  theme_bw()
```

## H2.4 
Participants with higher race IAT scores will give lower ratings to black applicants than white applicants. 

New variable defined as:\
`race_rating_diff` = mean rating on White applicants - mean rating on Black applicants\
\

Result: \
For linear regression race_rating_diff ~ IAT_Score_Race, slope = 0.090419, p-value= <2e-16. Higher IAT score, give higher ratings to white applicants than black applicants\

```{r}
c4 <- h2 %>%
  group_by(user_id, RACE) %>%
  summarise(race_mean_rating=mean(norm_rating_before)) 


for (i in 1:length(h2$user_id)) {
  user = h2$user_id[i]
  h2[h2$user_id==user, 'race_rating_diff'] <- c4[c4$user_id==user & c4$RACE=='White', ]$race_mean_rating - c4[c4$user_id==user & c4$RACE=='Black', ]$race_mean_rating
}
```


```{r}
mod.6 <- lm(race_rating_diff ~ IAT_Score_Race, h2, na.action=na.omit)
summary(mod.6)
```

```{r}
h2 %>%
  ggplot(aes(x=IAT_Score_Race,y=race_rating_diff))+
  geom_point()+
  geom_smooth(method='lm')+
  theme_bw()
```









